Imputation of Incomplete Multilevel Data with \pkg{mice}

### 2022-02-04 Manuscript/Manuscript.Rmd
Add popularity data v1 to conform to popmis data from mice

```{r pop, echo=FALSE, message=FALSE, warning=FALSE}
# randomly select a few classes
clusters <- sample(1:100, 10)

# load data 
pop <- readRDS("../Data/popNCR2.RDS") %>% 
  filter(class %in% clusters) %>% 
  mutate(class = as.factor(class))

# ICCs
icc_pop <- icc(popular ~ class, data = pop)
icc_teach <- icc(popteach ~ class, data = pop)
```

`popNCR2` is a simulated dataset with pupils clustered in classes, where the number of units $n = 2000$, and the number of clusters $N = 100$, on 7 variables:

  - `pupil`	Pupil number within class,
  - `class`	Class number,
  - `extrav`	Pupil extraversion,
  - `sex`	Pupil gender,
  - `texp`	Teacher experience (years),
  - `popular`	Pupil popularity,
  - `popteach`	Teacher popularity.

@@

```{r pop_pat, echo=FALSE, fig.cap = "Missing data pattern in the popularity data", message=FALSE, warning=FALSE}
# missingness
plot_md_pat(pop)
# TODO: add pattern by cluster, text mining example square size
# TODO: try to make it a shepley plot 
```

The popularity data is created such that there are strong relations between the incomplete variables and the clustering variable `class`. We can express this using the intra-class correlation (ICC). For `popular` the ICC is `r round(icc_pop, 2)`. For `popteach` it is `r round(icc_teach, 2)`. It would thus be wise to use multilevel modeling.

The missingness in this dataset is induced conform MAR and MNAR mechanisms. The missing data pattern, Figure \ref{fig:pop_pat}, shows that just one variable is incomplete **[the next part is not yet updated to reflect this]**.

To develop the best imputation model, we need to know whether the missingness in one variable depends on the observed values of other variables. Visual inspection usually suffices. We'll highlight only two variables to illustrate, but ideally one would inspect all relations. The questions we'll ask are: 'Does the missing data of `popular` depend on `popteach`?' and 'Does the missingness in teacher popularity depend on pupil popularity?' We'll evaluate this by making a histogram of `popteach` separately for the pupils with known popularity and missing popularity, and the other way around.

In Figure \ref{fig:pop_dist} we see that the distribution for the missing `popular` is further to the right than the distribution for observed `popular`. This would indicate a right-tailed MAR missingness. In fact, this is exactly what happens, because the missingness in these data was created manually. Now, we've made it observable by examining the relations between the missingness in popular and the observed data in `popteach`. There is also a dependency between the missingness in teacher popularity and pupil popularity. The relation seems to be right-tailed as well.

```{r pop_dist, fig.cap = "Conditional distributions in the popularity data", echo=FALSE, message=FALSE, warning=FALSE}
# plot_conditional(pop, x = "popteach", z = "popular", cluster = "class")
# TODO: check if up-side-down plot works
# TODO: check smoothing in geom_density function and make it the inverse of the sample size
# TODO: add functions to mice
# TODO: add title and informative legend
# TODO: add facets for some clusters
# or add propensity score distribution `is.na(popular) ~ .`
# TODO: make it average of cluster dens, not marginal
# alternatively add a quartile line around the density with geom_ribbon

## comments 24/1
# pop obs and pop miss in legend
# add to caption
# add fill
# overlap coeff??
# make densities proportional
# 
```

@@

```{r pop_ignored, echo=TRUE, message=FALSE, warning=FALSE}
# dry run to get imputation parameters
ini <- mice(pop, maxit = 0)

@@

# impute the data, ignoring the cluster structure
imp_ignored <- mice(pop, maxit = 1, pred = pred, print = FALSE)
```

@@

# # compare descriptives before and after imputation
# psych::describe(pop)[, c("n", "mean", "median", "min", "max", "sd")]
# psych::describe(mice::complete(imp_ignored))[, c("n", "mean", "median", "min", "max", "sd")] #note that this is just 1 imputation, not the pooled results
# TODO: add stripplot with boxplot overlay instead of the tables (make pooled one thick on top)
# TODO: pool mean median and sd
# feedback Stef: numbers, continuous statistics such as means, and uncertainty estimates. So we can pool the sd's. And leave out the min and max, because those are not normally distr.
# TODO: add FMI for each of the estimates? at least for the mean

# further inspection of the imputations
plot_box(imp_ignored, x = "popular", strip = TRUE)
# TODO: think about multimodality

# compare ICCs before and after imputation
ICCs <- data.frame(
  vars = c("popular", "popteach", "texp"), 
  CCA = c(icc(popular ~ class, pop), 
               icc(popteach ~ class, pop),
               icc(texp ~ class, pop)), 
  ignored = c(icc(popular ~ class, complete(imp_ignored)), 
              icc(popteach ~ class, complete(imp_ignored)), 
              icc(texp ~ class, complete(imp_ignored)))
  )
ICCs

@@

# impute the data, cluster as predictor
imp_predictor <- mice(pop, maxit = 1, pred = pred, print = FALSE)

@@

```{r pop_predictor_eval, echo=FALSE, message=FALSE, warning=FALSE}
# # check logged events
# head(imp_predictor$loggedEvents)
## "The mice() function detects multicollinearity, and solves the problem by removing one or more predictors for the model", in this case texp is removed as predictor of popular and popteach.

@@

# # compare descriptives before and after imputation
# psych::describe(pop)[, c("n", "mean", "median", "min", "max", "sd")]
# psych::describe(mice::complete(imp_predictor))[, c("n", "mean", "median", "min", "max", "sd")] #note that this is just 1 imputation, not the pooled results

# further inspection of the imputations
plot_box(imp_predictor, x = "popular", strip = TRUE)

# compare ICCs before and after imputation
ICCs <- ICCs %>% mutate(
           predictor = c(icc(popular ~ class, complete(imp_predictor)), 
                        icc(popteach ~ class, complete(imp_predictor)), 
                        icc(texp ~ class, complete(imp_predictor)))
           )
ICCs
```

@@

```{r pop_norm}
pred <- ini$pred
pred["popular", ] <- c(0, -2, 2, 2, 2, 0, 2) 
#-2 for the cluster variable, 2 for random effects
meth <- ini$meth
meth <- c("", "", "", "", "", "2l.norm", "")
imp_norm_2l <-
  mice(
    pop %>% mutate(class = as.integer(class)),
    pred = pred,
    meth = meth,
    maxit = 1,
    print = FALSE
  )
```

```{r pop_norm_eval, echo=FALSE}
# plot(imp_norm)
plot_box(imp_norm_2l, x = "popular", strip = TRUE)

ICCs <- ICCs %>% mutate(
           norm = c(icc(popular ~ as.factor(class), complete(imp_norm_2l)), 
                    icc(popteach ~ as.factor(class), complete(imp_norm_2l)), 
                    icc(texp ~ as.factor(class), complete(imp_norm_2l)))
           )
ICCs
```

@@

```{r pop_pan}
pred["popular", ] <- c(0, -2, 2, 2, 1, 0, 2)
meth <- c("", "", "", "", "", "2l.pan", "")
imp_pan_2l <-
  mice(
    pop %>% mutate(class = as.integer(class)),
    pred = pred,
    meth = meth,
    maxit = 1,
    print = FALSE
  )

@@

```{r pop_pan_eval, echo=FALSE}
# plot(imp_pan)
plot_box(imp_pan_2l, x = "popular", strip = TRUE)
ICCs <- ICCs %>% mutate(
           pan = c(icc(popular ~ as.factor(class), complete(imp_pan_2l)), 
                    icc(popteach ~ as.factor(class), complete(imp_pan_2l)), 
                    icc(texp ~ as.factor(class), complete(imp_pan_2l)))
           )
ICCs
```

>>>>

```{r pop, echo=FALSE, message=FALSE, warning=FALSE}
# randomly select a few classes
# clusters <- sample(1:100, 10)

# load data 
popmis <- mice::popmis %>% mutate(
  class = factor(school), 
  .keep = "unused")#readRDS("../Data/popNCR2.RDS") %>% 
  #filter(class %in% clusters) %>% 
  #mutate(class = as.factor(class))
popcomp <- foreign::read.spss("../Data/popular2.sav", to.data.frame = TRUE)
```

In this section we'll go over the different steps involved with imputing incomplete multilevel data. The data we're using is the `popmis` dataset from the `mice` package. This is a simulated dataset with pupils clustered in classes, where the number of units $n = 2000$, and the number of clusters $N = 100$, on 7 variables:

  - `pupil`	Pupil number within class,
  - `class`	Class number,
  - `extrav`	Pupil extraversion,
  - `sex`	Pupil gender,
  - `texp`	Teacher experience (years),
  - `popular`	Pupil popularity,
  - `teachpop`	Teacher popularity.

The analysis model corresponding to this dataset is multilevel regression with `popular` as the outcome variable, `sex` and `extrav` as unit-level predictors, and `texp` and `teachpop` as cluster-level predictors. Since the data is simulated and the missingness is induced, we can compare our imputations to the true data generating mechanism.

The data is created in such a way that the clustering variable `class` explains quite some variance in the outcome variable `popular`. We express this using the intra-class correlation,  ICC $=$ `r round(icc(popular ~ as.factor(class), data = popcomp), 2)`. This tells us that it would be wise to use multilevel modeling.

@@

```{r pop_pat, echo=FALSE, fig.cap = "Missing data pattern in the popularity data", message=FALSE, warning=FALSE}
# missingness
plot_md_pat(popmis)
# TODO: add pattern by cluster, text mining example square size
# TODO: try to make it a shepley plot 
```

The missingness is univariate and sporadic, which is illustrated in the missing data pattern in Figure \ref{fig:pop_pat}. The ICC in the incomplete data is `r round(icc(popular ~ as.factor(class), data = na.omit(popmis)), 2)`.

To develop the best imputation model for the incomplete variable `popular`, we need to know whether the missingness depends on the observed values of other variables. We'll highlight only one other variable to illustrate, but ideally one would inspect all relations. The questions we'll ask are: 'Does the missing data of pupil popularity (`popular`) depend on observed teacher popularity (`teachpop`)?'. This can be evaluated statistically, but visual inspection usually suffices. We'll make a histogram of `teachpop` separately for the pupils with known popularity and missing popularity.

In Figure \ref{fig:pop_dist} we see that **[update this part]** the distribution for the missing `popular` is further to the right than the distribution for observed `popular`. This would indicate a right-tailed MAR missingness. (In fact, this is exactly what happens, because the missingness in these data was created manually.) We've made it observable by examining the relations between the missingness in popular and the observed data in `teachpop`. 
<!-- There is also a dependency between the missingness in teacher popularity and pupil popularity. The relation seems to be right-tailed as well. -->

```{r pop_dist, fig.cap = "Conditional distributions in the popularity data", echo=FALSE, message=FALSE, warning=FALSE}
# popmis %>% 
#   filter(as.numeric(class) < 5) %>% 
#   ggplot(aes(x = teachpop, color = class)) +
#     geom_density() +
#     facet_wrap(~factor(is.na(popular), labels = c("popular missing", "popular observed")), ncol = 1) +
#     # scale_color_manual(values = plot_col, name = "popular") +
#     theme_classic() # +
#     # theme(legend.position = "bottom")

plot_conditional(popmis, x = "teachpop", z = "popular", cluster = "class")
# TODO: check if up-side-down plot works
# TODO: check smoothing in geom_density function and make it the inverse of the sample size
# TODO: add functions to mice
# TODO: add title and informative legend
# TODO: add facets for some clusters
# or add propensity score distribution `is.na(popular) ~ .`
# TODO: make it average of cluster dens, not marginal
# alternatively add a quartile line around the density with geom_ribbon

## comments 24/1
# pop obs and pop miss in legend
# add to caption
# add fill
# overlap coeff??
# make densities proportional

@@

```{r pop_ignored, echo=TRUE, message=FALSE, warning=FALSE}
# dry run to get imputation parameters
ini <- mice(popmis, maxit = 0)

@@

# impute the data, ignoring the cluster structure
imp_ignored <- mice(popmis, maxit = 1, pred = pred, print = FALSE)
```

@@

# # compare descriptives before and after imputation
# psych::describe(popmis)[, c("n", "mean", "median", "min", "max", "sd")]
# psych::describe(mice::complete(imp_ignored))[, c("n", "mean", "median", "min", "max", "sd")] #note that this is just 1 imputation, not the pooled results
# TODO: add stripplot with boxplot overlay instead of the tables (make pooled one thick on top)
# TODO: pool mean median and sd
# feedback Stef: numbers, continuous statistics such as means, and uncertainty estimates. So we can pool the sd's. And leave out the min and max, because those are not normally distr.
# TODO: add FMI for each of the estimates? at least for the mean

# further inspection of the imputations
plot_imps(imp_ignored, type = "stripplot", x = "popular")
# TODO: think about multimodality

# compare ICCs before and after imputation
ICCs <- data.frame(
  vars = c("popular", "teachpop", "texp"), 
  CCA = c(icc(popular ~ class, popmis), 
               icc(teachpop ~ class, popmis),
               icc(texp ~ class, popmis)), 
  ignored = c(icc(popular ~ class, complete(imp_ignored)), 
              icc(teachpop ~ class, complete(imp_ignored)), 
              icc(texp ~ class, complete(imp_ignored)))
  )
ICCs

@@

# impute the data, cluster as predictor
imp_predictor <- mice(popmis, maxit = 1, pred = pred, print = FALSE)

@@

```{r pop_predictor_eval, echo=FALSE, message=FALSE, warning=FALSE}
# # check logged events
# head(imp_predictor$loggedEvents)
## "The mice() function detects multicollinearity, and solves the problem by removing one or more predictors for the model", in this case texp is removed as predictor of popular and teachpop.

@@

# # compare descriptives before and after imputation
# psych::describe(popmis)[, c("n", "mean", "median", "min", "max", "sd")]
# psych::describe(mice::complete(imp_predictor))[, c("n", "mean", "median", "min", "max", "sd")] #note that this is just 1 imputation, not the pooled results

# further inspection of the imputations
plot_imps(imp_predictor, x = "popular", type = "stripplot")

# compare ICCs before and after imputation
ICCs <- ICCs %>% mutate(
           predictor = c(icc(popular ~ class, complete(imp_predictor)), 
                        icc(teachpop ~ class, complete(imp_predictor)), 
                        icc(texp ~ class, complete(imp_predictor)))
           )
ICCs
```

@@

```{r pop_norm}
pred <- ini$pred
pred["popular", ] <- c(0, -2, 2, 2, 2, 0, 2) 
#-2 for the cluster variable, 2 for random effects
meth <- ini$meth
meth <- c("", "", "", "", "", "2l.norm", "")
imp_norm_2l <-
  mice(
    popmis %>% mutate(class = as.integer(class)),
    pred = pred,
    meth = meth,
    maxit = 1,
    print = FALSE
  )
```

```{r pop_norm_eval, echo=FALSE}
# plot(imp_norm)
plot_imps(imp_norm_2l, x = "popular", type = "stripplot")

ICCs <- ICCs %>% mutate(
           norm = c(icc(popular ~ as.factor(class), complete(imp_norm_2l)), 
                    icc(teachpop ~ as.factor(class), complete(imp_norm_2l)), 
                    icc(texp ~ as.factor(class), complete(imp_norm_2l)))
           )
ICCs
```

@@

```{r pop_pan}
pred["popular", ] <- c(0, -2, 2, 2, 1, 0, 2)
meth <- c("", "", "", "", "", "2l.pan", "")
imp_pan_2l <-
  mice(
    popmis %>% mutate(class = as.integer(class)),
    pred = pred,
    meth = meth,
    maxit = 1,
    print = FALSE
  )

@@

```{r pop_pan_eval, echo=FALSE}
# plot(imp_pan)
plot_imps(imp_pan_2l, x = "popular", type = "stripplot")
ICCs <- ICCs %>% mutate(
           pan = c(icc(popular ~ as.factor(class), complete(imp_pan_2l)), 
                    icc(teachpop ~ as.factor(class), complete(imp_pan_2l)), 
                    icc(texp ~ as.factor(class), complete(imp_pan_2l)))
           )
ICCs
```

### 2022-02-04 Manuscript/Manuscript.tex
Add popularity data v1 to conform to popmis data from mice

\texttt{popNCR2} is a simulated dataset with pupils clustered in
classes, where the number of units \(n = 2000\), and the number of
clusters \(N = 100\), on 7 variables:

\begin{itemize}
\tightlist
\item
  \texttt{pupil} Pupil number within class,
\item
  \texttt{class} Class number,
\item
  \texttt{extrav} Pupil extraversion,
\item
  \texttt{sex} Pupil gender,
\item
  \texttt{texp} Teacher experience (years),
\item
  \texttt{popular} Pupil popularity,
\item
  \texttt{popteach} Teacher popularity.
\end{itemize}

@@

The popularity data is created such that there are strong relations
between the incomplete variables and the clustering variable
\texttt{class}. We can express this using the intra-class correlation
(ICC). For \texttt{popular} the ICC is 0.4. For \texttt{popteach} it is
0.36. It would thus be wise to use multilevel modeling.

The missingness in this dataset is induced conform MAR and MNAR
mechanisms. The missing data pattern, Figure \ref{fig:pop_pat}, shows
that just one variable is incomplete \textbf{{[}the next part is not yet
updated to reflect this{]}}.

To develop the best imputation model, we need to know whether the
missingness in one variable depends on the observed values of other
variables. Visual inspection usually suffices. We'll highlight only two
variables to illustrate, but ideally one would inspect all relations.
The questions we'll ask are: `Does the missing data of \texttt{popular}
depend on \texttt{popteach}?' and `Does the missingness in teacher
popularity depend on pupil popularity?' We'll evaluate this by making a
histogram of \texttt{popteach} separately for the pupils with known
popularity and missing popularity, and the other way around.

In Figure \ref{fig:pop_dist} we see that the distribution for the
missing \texttt{popular} is further to the right than the distribution
for observed \texttt{popular}. This would indicate a right-tailed MAR
missingness. In fact, this is exactly what happens, because the
missingness in these data was created manually. Now, we've made it
observable by examining the relations between the missingness in popular
and the observed data in \texttt{popteach}. There is also a dependency
between the missingness in teacher popularity and pupil popularity. The
relation seems to be right-tailed as well.

@@

\begin{CodeChunk}
\begin{CodeInput}
R> # dry run to get imputation parameters
R> ini <- mice(pop, maxit = 0)
R> 
R> # extract predictor matrix and adjust
R> pred <- ini$pred
R> pred[, c("class", "pupil")] <- 0
R> 
R> # impute the data, ignoring the cluster structure
R> imp_ignored <- mice(pop, maxit = 1, pred = pred, print = FALSE)
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeOutput}
      vars       CCA   ignored
1  popular 0.3989463 0.4235643
2 popteach 0.3605902 0.3605902
3     texp 1.0000000 1.0000000
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> # adjust the predictor matrix
R> pred <- ini$pred 
R> pred[, "pupil"] <- 0
R> 
R> # impute the data, cluster as predictor
R> imp_predictor <- mice(pop, maxit = 1, pred = pred, print = FALSE)
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeOutput}
      vars       CCA   ignored predictor
1  popular 0.3989463 0.4235643 0.3904372
2 popteach 0.3605902 0.3605902 0.3605902
3     texp 1.0000000 1.0000000 1.0000000
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> pred <- ini$pred
R> pred["popular", ] <- c(0, -2, 2, 2, 2, 0, 2) 
R> #-2 for the cluster variable, 2 for random effects
R> meth <- ini$meth
R> meth <- c("", "", "", "", "", "2l.norm", "")
R> imp_norm_2l <-
+   mice(
+     pop %>% mutate(class = as.integer(class)),
+     pred = pred,
+     meth = meth,
+     maxit = 1,
+     print = FALSE
+   )
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeOutput}
      vars       CCA   ignored predictor      norm
1  popular 0.3989463 0.4235643 0.3904372 0.4165006
2 popteach 0.3605902 0.3605902 0.3605902 0.3605902
3     texp 1.0000000 1.0000000 1.0000000 1.0000000
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> pred["popular", ] <- c(0, -2, 2, 2, 1, 0, 2)
R> meth <- c("", "", "", "", "", "2l.pan", "")
R> imp_pan_2l <-
+   mice(
+     pop %>% mutate(class = as.integer(class)),
+     pred = pred,
+     meth = meth,
+     maxit = 1,
+     print = FALSE
+   )
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeOutput}
      vars       CCA   ignored predictor      norm       pan
1  popular 0.3989463 0.4235643 0.3904372 0.4165006 0.3827595
2 popteach 0.3605902 0.3605902 0.3605902 0.3605902 0.3605902
3     texp 1.0000000 1.0000000 1.0000000 1.0000000 1.0000000
\end{CodeOutput}
\end{CodeChunk}

>>>>

In this section we'll go over the different steps involved with imputing
incomplete multilevel data. The data we're using is the \texttt{popmis}
dataset from the \texttt{mice} package. This is a simulated dataset with
pupils clustered in classes, where the number of units \(n = 2000\), and
the number of clusters \(N = 100\), on 7 variables:

\begin{itemize}
\tightlist
\item
  \texttt{pupil} Pupil number within class,
\item
  \texttt{class} Class number,
\item
  \texttt{extrav} Pupil extraversion,
\item
  \texttt{sex} Pupil gender,
\item
  \texttt{texp} Teacher experience (years),
\item
  \texttt{popular} Pupil popularity,
\item
  \texttt{teachpop} Teacher popularity.
\end{itemize}

The analysis model corresponding to this dataset is multilevel
regression with \texttt{popular} as the outcome variable, \texttt{sex}
and \texttt{extrav} as unit-level predictors, and \texttt{texp} and
\texttt{teachpop} as cluster-level predictors. Since the data is
simulated and the missingness is induced, we can compare our imputations
to the true data generating mechanism.

The data is created in such a way that the clustering variable
\texttt{class} explains quite some variance in the outcome variable
\texttt{popular}. We express this using the intra-class correlation, ICC
\(=\) 0.36. This tells us that it would be wise to use multilevel
modeling.

@@

The missingness is univariate and sporadic, which is illustrated in the
missing data pattern in Figure \ref{fig:pop_pat}. The ICC in the
incomplete data is 0.56.

To develop the best imputation model for the incomplete variable
\texttt{popular}, we need to know whether the missingness depends on the
observed values of other variables. We'll highlight only one other
variable to illustrate, but ideally one would inspect all relations. The
questions we'll ask are: `Does the missing data of pupil popularity
(\texttt{popular}) depend on observed teacher popularity
(\texttt{teachpop})?'. This can be evaluated statistically, but visual
inspection usually suffices. We'll make a histogram of \texttt{teachpop}
separately for the pupils with known popularity and missing popularity.

In Figure \ref{fig:pop_dist} we see that \textbf{{[}update this part{]}}
the distribution for the missing \texttt{popular} is further to the
right than the distribution for observed \texttt{popular}. This would
indicate a right-tailed MAR missingness. (In fact, this is exactly what
happens, because the missingness in these data was created manually.)
We've made it observable by examining the relations between the
missingness in popular and the observed data in \texttt{teachpop}.

{\centering \includegraphics{Manuscript_files/figure-latex/pop_dist-1}

\caption[Conditional distributions in the popularity data]{Conditional distributions in the popularity data}\label{fig:pop_dist}
\end{figure}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> # dry run to get imputation parameters
R> ini <- mice(popmis, maxit = 0)
R> 
R> # extract predictor matrix and adjust
R> pred <- ini$pred
R> pred[, c("class", "pupil")] <- 0
R> 
R> # impute the data, ignoring the cluster structure
R> imp_ignored <- mice(popmis, maxit = 1, pred = pred, print = FALSE)
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeOutput}
      vars       CCA   ignored
1  popular 0.5583543 0.4739527
2 teachpop 0.6823380 0.6823380
3     texp 1.0000000 1.0000000
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> # adjust the predictor matrix
R> pred <- ini$pred 
R> pred[, "pupil"] <- 0
R> 
R> # impute the data, cluster as predictor
R> imp_predictor <- mice(popmis, maxit = 1, pred = pred, print = FALSE)
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeOutput}
      vars       CCA   ignored predictor
1  popular 0.5583543 0.4739527 0.5660108
2 teachpop 0.6823380 0.6823380 0.6823380
3     texp 1.0000000 1.0000000 1.0000000
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> pred <- ini$pred
R> pred["popular", ] <- c(0, -2, 2, 2, 2, 0, 2) 
R> #-2 for the cluster variable, 2 for random effects
R> meth <- ini$meth
R> meth <- c("", "", "", "", "", "2l.norm", "")
R> imp_norm_2l <-
+   mice(
+     popmis %>% mutate(class = as.integer(class)),
+     pred = pred,
+     meth = meth,
+     maxit = 1,
+     print = FALSE
+   )
\end{CodeInput}
\begin{CodeOutput}
Warning: Number of logged events: 1
\end{CodeOutput}
\end{CodeChunk}

\begin{CodeChunk}
\begin{CodeOutput}
Warning: Removed 5088 rows containing non-finite values (stat_boxplot).
\end{CodeOutput}
\begin{CodeOutput}
Warning: Removed 5088 rows containing missing values (geom_point).
\end{CodeOutput}

@@

\begin{CodeOutput}
      vars       CCA   ignored predictor      norm
1  popular 0.5583543 0.4739527 0.5660108 0.5583543
2 teachpop 0.6823380 0.6823380 0.6823380 0.6823380
3     texp 1.0000000 1.0000000 1.0000000 1.0000000
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> pred["popular", ] <- c(0, -2, 2, 2, 1, 0, 2)
R> meth <- c("", "", "", "", "", "2l.pan", "")
R> imp_pan_2l <-
+   mice(
+     popmis %>% mutate(class = as.integer(class)),
+     pred = pred,
+     meth = meth,
+     maxit = 1,
+     print = FALSE
+   )
\end{CodeInput}
\begin{CodeOutput}
Warning: Number of logged events: 1
\end{CodeOutput}
\end{CodeChunk}

\begin{CodeChunk}
\begin{CodeOutput}
Warning: Removed 5088 rows containing non-finite values (stat_boxplot).
\end{CodeOutput}
\begin{CodeOutput}
Warning: Removed 5088 rows containing missing values (geom_point).
\end{CodeOutput}

@@

\begin{CodeOutput}
      vars       CCA   ignored predictor      norm       pan
1  popular 0.5583543 0.4739527 0.5660108 0.5583543 0.5583543
2 teachpop 0.6823380 0.6823380 0.6823380 0.6823380 0.6823380
3     texp 1.0000000 1.0000000 1.0000000 1.0000000 1.0000000
\end{CodeOutput}
\end{CodeChunk}
